CSD, Disjoint components and trayectory detection

### 2019-03-26 ArticuloCSDyCM/Paper03.tex
mas paiper

Spatially tracking succesive activity of electro-physiological data in a concise,
rigorous and interpretable manner poses a challenging problem. Measurement recordings
are usually presented first as a series of traces that correspond each to a sampling
point on the tissue. As Micro Electrode Arrays (MEAs) become denser, this representation
of the data becomes saturated and the discovery of patterns turns into a difficult task.
Novel information is obfuscated by the same high density acquisition that allows us to
discover it. Representing  the recorded data in ways which can unveil previously
invisible patterns is a necessity. Various techniques for separation of functional
generators are available now, most derived from the application of blind source
separation analysis, such as ICA. 
Our task in this paper is to present a method for tracing out
``functional local connectivity'', which would bring independent information
that is complementary to those analysis.  Our  method would be applicable mostly to
activity in highly structured neural tissue. We shall exemplify it by application to
the data obtained from recordings of the rats hippocampus, taking advantage of its lammiar orderer structure.

@@

Our method permits us to map possible trajectories of activity inside structured neuronal tissue, without requiring ad hoc geometric constraints. As a prime example, non-convex structures which may show simultaneous activity in separate parts could show the robustness of our method. We have tested then the method on slices of rat's hippocampus, concentrating in the CA region. We exemplify the usage of the method with data that has been obtained by an experiment with ...\textbf{bla bla bla describr experimentos}.

@@

After aquiring data from experiment, we pass the data through some preprocessing
routines. For the most part, these serve to detect and discard unusable channels,
such as saturated or failling electrodes. These vary from data set to data set,
as they can be produced by bubbles or by saturation after electrical stimulation.
We call this step the elimination of bad channels. We use a simple criterion for this:
if the absolute value of the signal is above certain threshold (wich must be
adjusted for evocated vs spontaneous acttivity) or if the standar deviation is
larger than certain heuristically determined value, the channel is discarded.
A similar but orthogonal criterium can be used for detection of channels with
spikes.

@@

The concepts presented here require first obtaining the CSD from the recorded LFP. Any method would present the same utility. The first method is a  finite difference operator, while the second is the kCSDA presented by Potworowski \emph{et all} \cite{Potworowski2011}, which is an inverse problem approach.  Where both methods are usable, we prefer the former, as is faster and requires less assumptions. We have used both methods, kCSDA where
the difference method is not applicable. The results are
independent of the method save for multiplicative factor, which
are irrelevant in the Center of Mass tracking.

Thanks to the density and scales of the BioCAM 4096, a numerical finite difference method
can give a good approximation to the CSD, if the number of failing channels is low or
are outside the r.o.i.
For experiments that were made without electrical stimulation this turned out to be true, with the order of saturated or otherwise failing electrodes in the tens.  A  numerical Laplacian Operator which reduces the cross effect of rectangular grids is the convolution of the data with the following matrix \cite{Lindberg90}:
\begin{equation}
\nabla^2_{1/3}=2/3
\begin{pmatrix}
  0 & 1 & 0 \\
  1 & -4 & 1 \\
  0 & 1 & 0
\end{pmatrix}
+1/3
\begin{pmatrix}
  0.5 & 0 & 0.5 \\
  0 & -2 & 0 \\
  0.5 & 0 & 0.5
\end{pmatrix}  
\end{equation}
This sort of operators are extremely sensitive to abrupt changes in the values. The noise of the data can be perceived in the spatial domain as rapidly varying edges. A Gaussian spatial denoising can be performed before or after the Laplacian operator to the data in order to reduce excessive borders. Our Gaussian blur filter has a $\sigma$ value of three electrodes or $126 \mu m$. Intuitively this means that it smooths over each structure smaller than the soma of a typical pyramidal cell. 
This is  in accordance to our mean field approach, where we expect smooth differentiable fields over the scale of the neurons, but not in finer scales (for details see \cite{Bedard11}) .

On previous works it has been argued that numerical difference operators sacrifice all data on the borders of the array. For electrode arrays of lesser density, this could posse a problem, but in our case we only renounce to 27 out of 4095 channels (one is grounded), and those are outside the region of interest.
If very few electrodes fail, more sophisticated methods such as iCSDA \cite{Leski2011} and kCSDA \cite{Potworowski2011} prove little advantage in estimating the sources under the ohmic, isotropic and homogeneous assumptions. In case that around 30 or more electrodes
fail, specially if they are on the region of interest, the use of such methods
becomes necessary. The kCSDA method relies on the
assumption that a finite family of convenient functions can span, with acceptable accuracy, the possible CSD at each instant in time. The idea behind the method is simple:
One presents a model function of this family and obtains the LFP that such a function
would generate. Then one calculates the coefficients for the entire family by projecting the
experimental data over this ideal LFP model function family. A general presentation is in the reference \cite{Potworowski2011},we shall recount only the basic implementation that we used.

We shall borrow the notation in the original paper \cite{Potworowski2011}. Our model function for the CSD is a hard disk function, described by:
\begin{equation}
  \tilde{b}_j(x,y)=\begin{cases}
  1, \text{ for } (x-x_j)^2+(y-y_j)^2 \leq R^2 \\
  0 \text{ otherwise.} 
  \end{cases}
\end{equation}
The $R$ parameter is the effective radius of the charge density, which for our
calculations was taken as half the inter-electrode distance, that is $R=21 \mu m$.
Each pair $(x_j, y_j)$ is the center of one such model function, which we take as
the position of each  functioning electrode in an adequate coordinate system.
The LFP generated by this function would then be:
\begin{equation}
  \tilde{b}_j(x,y)=\frac{1}{2\pi\sigma}
  \iint \! \rd x \rd y \arcsinh \biggr(
  \frac{2 h } {\sqrt{(x-x_j)^2+(y-y_j)^2}}
  \biggl)
  \tilde{b}_j(x,y),
\end{equation}
  with $h$ being the effective height of the function support, in our case,
  the half width of the hippocampus slice ($h=150 \mu m$).

Then the kCSDA consist of projecting the experimental data (LFP), frame by frame,
over these model functions and obtaining the  coefficients that would spawn an idealized
LFP approximation. These coefficients are the same that would produce the
corresponding CSD. The first operator is the projection operator $K$ of the experimental
LFP (from now on $V(x,y)$) on the space of the idealized LFP (which we call $V^* (x,y)$).
\begin{equation}
  K_{jk}=\sum_{l=1}^{M}b_l(x_j,y_j)b_l(x_k,y_k).
\end{equation}
The sum is done over the set of $M$ electrodes that we want to use.
The other operator is the one that maps the space of $V^*(x,y)$ into the idealized
CSD functions, now called $C^*(x,y)$. 
\begin{equation}
  \tilde{K}_{jk}=\sum_{l=1}^Mb_l(x_j,y_j)\tilde{b}_l(x_k,y_k).
\end{equation}
Then $C^*(x,y)$ can be obtained on the region of interest by:
\begin{equation}
  C^*(x_j,y_j)=\sum_{k,l=1}^M \tilde{K}_{kj} K^{-1}_{ml} V(x_l,y_l)
\end{equation}
Notice how the method incorporates a denoising mechanism, so
no smoothing procedures are used.
The great advantage of this method is that is possible to obtain a reasonable
approximation to CSD even when the presence of unusable channels is near or inside
the region of interest. The difference method, on the other hand, affects all surrounding
channels to a failing channel, up to 8 of them, so it becomes unusable if enough
saturated electrodes are over the r.o.i.

@@

The attainment of $K$ and $\tilde{K}$ turns out to be computer-costly for high density MEAs and dense temporal sampling, but paraletization, specially on
GPUs, solves this issue.
We have performed comparison of both methods over experimental data that has
no failing channels in order to adjust the parameters of the second method. We
have obtained consistent results between them.

@@

Center of Mass Analysis (CMA) has been used previously in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Chao07, Manjarrez07, Manjarrez09}. In such works the chosen ``mass'' was an heuristic measure adecuate for representing the corse-grain picture of displacement of activity. These present the problem of justifing the choice made on practical grounds. For calculating a vectorial average, the adecuate wheight per vector is a density, a positive quantity that measures the concentration of something per space unit, so it is only adequate that we search for a natural density in the electrophysiological experiments. The Current Source Density fits the mathematical and interpretative requirements. We shall denote it in the following as $C$ (following \cite{Potworowski2011}).  This measure is usually presented in arbitrary units, which permits us to interpret it two-fold. One, directly from its definition, as the \emph{differential divergence} of the vectorial electric current density:
\begin{equation}
  C:=\nabla \cdot J
\end{equation}
Due to conservation of charges, this is the same as as the temporal variation of density of charge:
\begin{equation}
  C=\frac{\partial \rho}{\partial t}
\end{equation}
Current Source Density as a mathematical density (a quasi-probability density function)  allows  to rigorously apply the concepts of vector averaging. This use has been overlooked. This might have occurred because  samplings at lesser spatial resolution did not permit estimations of this quantity with enough precision to make reliable spatial implications, even with the use of sophisticated CSD estimation methods. In our case, we want to use the density as a mathematician would: a function which assigns weights to points in the space of interest and permits us to carry linear operations over them.  The space here is the sampling points of the electrodes at a given instant in time, and the weight, the CSD, is amount of interchange of charged ions from the inside to the outside of cellular membranes, per unit space.  This is an indicator of activity in the neurons, therefore the CSD is also, indirectly, a \emph{``density of local activity''}.

Apparently, CSD would not fit the requirements of non-negativity for its use as a mathematical density, but this turns out to be not a problem.  CSD separates the measured activity into three distinct sets, namely, the set of all sources, the set of all sinks and the zero set. The separation is not quantitative but qualitative: even when the signs of electric charge are arbitrary, the distinction between them is not. Therefore, CSD represents on each of the two active sets (sinks and sources) exactly what we need, a unit of concentration per unit space, in this case, per area, and the sign can be ignored, as long as we perform the operations in each set separately. Also, the separation by the zero value set of the CSD is in contrast to the zero value of the LFP, where it has no precise meaning as it only represents a practical reference value. This is acknowledged in most of the literature, where only the scales of the LFP recording are shown, without reference to the sign of the values. In the CSD representation,  the so called zero set (all the points in our space which have a recorded CSD of zero)  has a precise meaning: it is an instantaneous lack of activity and a border between sinks and sources. This shall be exploited further in the section below.

Once the separation into sinks and sources has been made, we would have two densities, or more precisely, two different density functions defined over two separate sets. Trying to use this for obtaining vector averages would still be crude and inexact. If these sets spread out over large non-convex geometries, the vector averages could lie outside the sets, having little or none physiological interpretation. Therefore we decided to introduce another concept from geometrical sets in order to reduce the analysis to local effects.

\subsection{CSD and Disjoint Components}

A first visual inspection of the CSD color map at a given instant in time shows that both sources and sinks appear in rather large and significant patches. These appear to correspond to the expected double and triple poles of firing neurons in structured tissue \cite{Buzsaki2012} and inspired us to separate the sink and source  sets into their \emph{disjoint components}. This concept is used in mathematics to denote subsets of a given geometric set that do not touch each other or share no elements in common \cite{Halmos}. Their rough correspondence to poles of firing neurons would be most useful in tracking the successive locus of the activity.

To clarify the concept and how we are we using it an example is in order. In figure \ref{Disjuntos} A we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set $A$ consists of a single component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set $B$ is made of two disjoint components, labeled $B1$ and $B2$. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set $B$ consists of two disjoint components, but each component can be regarded as a connected set on their own. Likewise the set $C$ is made of three disjoint components. In the subfigure \ref{Disjuntos} Aiii we show the center of mass of the sets. The set $A$ is concave, so, it can have a center of mass outside the set. The
set $B$ has also a center of mass outside the set. It has two disjoint components that are aproximatelly
equall and separated by a long distance (this is not an implication: a set can be disjoint and have
a CM inside the set). For the set $C$ our aproach has been used: we treat every disjoint component of the
set as an independent entity and calculate the center of mass for each one. Notice that this doesn't preclude the posibility of having a concave subset with a CM that lies outside. But, as we shall see,
as the electrode resolution is finite, the posibility of having highly non compact shapes is unlikely, and they
tend to have a CM inside or near their borders. In the figure \ref{Disjuntos} B we show this with experimental
data (we are using evoked activity data, as explained in section \ref{sec:evocada}). In graph Bi we
sepparate the CSD data into the sink and sources set. We take sinks as example. In the graph Bii
we label each disjoint component, obtained by a single pass algorhythm. There are 25 components, and
the biggest two are not strictly convex but they have tolerable deviations from it. In the
subfigure Biii we indicate the CM of only those components that are above certain threshold of
integrated intensity (the dot is proportional in area to such intensity, there are five CM to be
seen). Each CM gets listed as a probable step in a trajectory.

\begin{figure}
  \includegraphics[width=0.9\textwidth]{Figuras/Disjuntos01.pdf}
  \caption{ Explanation of the geometrical concepts used on our analysis. In subfigure A we show
    with an artificial example the concepts of disjointness and convexivity. The subfigure Ai shows
    three sets, $A$, $B$, and $C$. $A$ fails to be convex because it is concave, and $C$ fails to be
    be convex because it is made of two disjoint components. In each of them is possible to draw a line
    connecting two points inside the set that is not entirely inside the set. In $B$ that is not possible, the set $B$ is convex in a mathematical sense. In subfigure Aii we made some modification to show that
    each set can be made of many disjoint components, and each one of them can be convex or not.  In Aiii we show the center of mass for $A$ and $B$ taken as a whole, while for $C$ we break the set into each of the disjoint components and calculate each CM separately. There is no guarantee that each disjoint component would be convex, but as the resolution of the data is finite, the smaller components are more likely to
    be convex. In figure B we shoe everything with a frame of experimental data. In Bi we sepparate the
    data into three sets, Sources, Sinks, and Zeros. We discard the Zeros and keep the interesting ones.
    In Figure Bii we take the Sink Set and separate it into disjoint components, that are labeled from one to 25 in this particular example. From this subsets we calculate the total mass of each one and only
    considerate for the next step those who are above certain threshold of integrated intensity. The
    next step is shown in Biii, where we plot the Center of Mass for each of such subsets. Deviations
    from convexity seem to be unimportant, as most of the mass is concentrated in the inner electrodes
    of each set.
  }
  \label{disjuntos}
\end{figure}

@@

The ``thick zero'' would help us to sepparate the sinks and sources sets into its disjoint components, which would correspond then to clear patches of definite activity. If each one of these patches could be interpreted as an active unit, we may asign to it a ``center of activity'' in order to follow their displacements. A Center of Mass for each instantaneous disjoint sink/source would provide us a pragmatic measurement of an ``active locus''. We can then calculate at each time frame the CM for each disjoint component before tracking displacements of activity.

At each disjoint component at a precise time $t$, indexed $k$, we have a number of points with CSD of the same sign, indexed $j$. Then, the CM of the $k$ component would have the usual definition:
\begin{equation}\label{cmparadisj}
   \langle q(t) \rangle_k =\frac{\sum_j q_{j,k} (t) I_{j,k} (q_{j,k},t)}
           {\sum_j I(q_{j,k},t)},
\end{equation}
where $I(q_{j,k},t)$ is the CSD at point $q_{j,k}=(x_{j,k}, y_{j,k})$ at time $t$. These disjoint components, while not rigorously convex, are less spread over the space of the measurements and tend to have their CM over the actual active points, or at the border of the active units.

@@

We must not take out of sight the phenomena which is producing the CSD distribution: neurons have a finite size, and the effect of their action a delimited extent. Every patch that we can identify as a disjoint component of the Sources/Sink sets indicates a  joint active set of units. A center of mass for each disjoint component would then indicate an average locus for this active group, a putative center of activity to which we can give a position, intensity and time. An instantaneous snapshot of such information would depict the centers of activity of an ordered structure of neurons and their relative intensities. But it is in the successive depiction of a series of such snapshots that this analysis can show interpretative power. The appearance , displacement, and disappearance of such centers could reveal very subtle details of the dynamics of neural activity. Very small, fast displacements of the Centers of mass, may be occur due to slight asynchronous between neurons which belong to the same active group. Larger displacements and  longer intervals (more than two or so firing periods) could indicate physiological connections between different groups.

The next step in our analysis is a procedure to systematically associate to each CM at time $t$ a ``successor'' at time $t+\Delta t$, where $\Delta t$ is the sampling time step. We used an euclididan distance. Let $q_1=(x_1, y_1)$ be the coordinates of the first point and $q_2=(x_2,y_2)$, the corresponding coordinates of the second. Then the distance is:
\begin{equation}
d(q_1,q_2):=\sqrt{(x_1-x_2)^2+(y_1+y_2)^2}.
\end{equation}
The integrated intensity of a CM is simply the denominator in the eq. \ref{cmparadisj}, that is, the total ``mass'' obtained by integrating the CSD in a specific disjoint component:
\begin{equation}
  I_k(t)=\sum_j I (q_{j,k}, t).
\end{equation}
We specify then a tolerance $\delta$. If a certain $t$ a CM is below $\delta$ distance from another at $t+\Delta t$, we consider the latter the successor in time of the former. By continuously applying such procedure, we can begin to trace trajectories of the CMs, as they appear, wander, and fade.

When a specific trajectory cannot find a successor point that is above the
threshold intensity, we assume that it is finished, so we do not concatenate
points that are separated in time by more than $\Delta t$. In order to paste different
trajectories, more rigorous analysis of their origin and interpretation would be needed.

@@

\subsection{Application to evoked activity in idem.}

@@

>>>>

Spatially tracking succesive activity of electro-physiological data in a concise,
rigorous and interpretable manner poses a challenging problem. Measurement recordings
are usually presented first as a series of traces that correspond each to a sampling
electrode on the tissue. Then a series of causality tests are made on the
signal to search for possible efective conections. 
As Micro Electrode Arrays (MEAs) become denser, that representation
of the data saturates itself  and the discovery of patterns becomes a difficult task.
Novel information is obfuscated by the same high density acquisition that allows us to
discover it. Representing  the recorded data in ways which can unveil previously
invisible patterns is a necessity. Various techniques for separation of functional
generators are available now, most derived from the application of blind source
separation analysis, such as ICA. 
Our task in this paper is to present a method for tracing out
``functional local connectivity'', which would bring independent information
that is complementary to those analysis.  Our  method would be applicable mostly to
activity in highly structured neural tissue. We shall exemplify it by application to
the data obtained from recordings of the rats hippocampus,
taking advantage of its lammiar orderer structure.

@@

Our method permits us to map possible trajectories of activity inside structured neuronal tissue, without requiring ad hoc geometric constraints. As a prime example, non-convex structures which may show simultaneous activity in separate parts could show the robustness of our method. We have tested then the method on slices of rat's hippocampus, concentrating in the CA region. We exemplify the usage of the method with data that has been obtained on experiments on rat's hippocampal slices that are centered on the GD and CA
regions. The slices are put over an array of 64 by 64 electrodes
(BioCAM 4096 by 3Brain \cite{BioCam}),
and data is recorded at sampling rates between 7 and 14 kHz. The richness of the
data permits us to obtain a prime use case for our method. Downscaling is
possible in principle, but the array of electrodes should be at least 2D and
a interelectrode distance of the order of magnitude of a typical pyramidal soma
(around 100 $\mu$m).

@@

After aquiring data from experiment, we pass the data through some preprocessing
routines. These serve to detect and discard unusable channels,
such as saturated or failling electrodes. These vary from data set to data set,
as they can be produced by bubbles or by saturation after electrical stimulation
or other random causes. 
We call this step the elimination of bad channels. We use a simple criterion for this:
if the absolute value of the signal is above certain threshold (wich must be
adjusted for evocated vs spontaneous acttivity) or if the standar deviation is
larger than certain heuristically determined value, the channel is discarded.

%% esto puede no ser necesario. 
A similar but orthogonal criterium can be used for detection of channels with
spikes.

@@

The concepts presented here require first obtaining the CSD from the recorded LFP. Any method would present the same utility, we have used two different
approaches. The first method is a  finite difference operator, while the second is the kCSDA presented by Potworowski \emph{et all} \cite{Potworowski2011}, which is an inverse problem approach.  Where both methods are usable, we prefer the former,
as it is faster and requires less assumptions.
We have used both methods, kCSDA where
the difference method is not applicable. The results are
independent of the method save a for multiplicative factor, which
is irrelevant in the Center of Mass tracking.

Thanks to the density and scales of the BioCAM 4096, a numerical finite difference method
produces an excelent approximation to the CSD,
if the number of failing channels is low or
are outside the r.o.i.
For experiments that were made without electrical stimulation is the case, with the order of mangitude of unusable channels in the tens.  A  numerical Laplacian Operator which reduces the cross effect of rectangular grids is the convolution of the data with the following matrix \cite{Lindberg90}:
\begin{equation}
\nabla^2_{1/3}=2/3
\begin{pmatrix}
  0 & 1 & 0 \\
  1 & -4 & 1 \\
  0 & 1 & 0
\end{pmatrix}
+1/3
\begin{pmatrix}
  0.5 & 0 & 0.5 \\
  0 & -2 & 0 \\
  0.5 & 0 & 0.5
\end{pmatrix}  
\end{equation}
This sort of operators are extremely sensitive to abrupt changes in the values. The noise of the data can be perceived in the spatial domain as rapidly varying edges. A Gaussian spatial denoising can be performed before or after the Laplacian operator to the data in order to reduce excessive borders. Our Gaussian blur filter has a $\sigma$ value of $126 \mu m$, which corresponds to three times the interelectrode distance.
Intuitively this means that it smooths over each structures smaller than the soma of a typical pyramidal cell. 
This is  in accordance to our mean field approach, where we expect smooth differentiable fields over the scale of the neurons, but not in finer scales (for details see \cite{Bedard11}).

On previous works it has been argued that numerical difference operators sacrifice all data on the borders of the array. For electrode arrays of lesser density, this could posse a problem, but in our case we only renounce to 27 out of 4095 channels (one is grounded), and those are outside the region of interest.
If very few electrodes fail, more sophisticated methods such as iCSDA \cite{Leski2011} and kCSDA \cite{Potworowski2011} prove little advantage in estimating the sources under the ohmic, isotropic and homogeneous assumptions. In case that around 30 or more electrodes
fail, specially if they are on the region of interest, the use of such methods
becomes necessary. The kCSDA method relies on the
assumption that a finite family of convenient functions can span, with acceptable accuracy, a very good aproximation to the  CSD at each instant in time.
 A general presentation is in the reference \cite{Potworowski2011},we shall recount only the basic implementation that we used.
One presents a family of model functions for the CSD and
obtains the LFP that such functions
would generate. Then one calculates the coefficients for the entire family by projecting the
experimental data over this ideal LFP model function family.
We shall borrow the notation in the original paper \cite{Potworowski2011}. Our model function for the CSD is a hard disk function, described by:
\begin{equation}
  \tilde{b}_j(x,y)=\begin{cases}
  1, \text{ for } (x-x_j)^2+(y-y_j)^2 \leq R^2 \\
  0 \text{ otherwise.} 
  \end{cases}
\end{equation}
The $R$ parameter is the effective radius of the charge density, which for our
calculations was taken as half the inter-electrode distance, that is $R=21 \mu m$.
Each pair $(x_j, y_j)$ is the center of one such model function, which we take as
the position of each  functioning electrode in an adequate coordinate system.
The LFP generated by this function would then be:
\begin{equation}
  \tilde{b}_j(x,y)=\frac{1}{2\pi\sigma}
  \iint \! \rd x \rd y \arcsinh \biggr(
  \frac{2 h } {\sqrt{(x-x_j)^2+(y-y_j)^2}}
  \biggl)
  \tilde{b}_j(x,y),
\end{equation}
  with $h$ being the effective height of the function support, in our case,
  the half width of the hippocampus slice ($h=150 \mu m$).

Then the kCSDA consist of projecting the experimental data (LFP), frame by frame,
over these model functions and obtaining the  coefficients that would spawn an idealized
LFP approximation. These coefficients are the same that would produce the
corresponding CSD. The first operator is the projection operator $K$ of the experimental
LFP (from now on $V(x,y)$) on the space of the idealized LFP (which we call $V^* (x,y)$).
\begin{equation}
  K_{jk}=\sum_{l=1}^{M}b_l(x_j,y_j)b_l(x_k,y_k).
\end{equation}
The sum is done over the set of $M$ electrodes that we want to use.
The other operator is the one that maps the space of $V^*(x,y)$ into the idealized
CSD functions, now called $C^*(x,y)$. 
\begin{equation}
  \tilde{K}_{jk}=\sum_{l=1}^Mb_l(x_j,y_j)\tilde{b}_l(x_k,y_k).
\end{equation}
Then $C^*(x,y)$ can be obtained on the region of interest by:
\begin{equation}
  C^*(x_j,y_j)=\sum_{k,l=1}^M \tilde{K}_{kj} K^{-1}_{ml} V(x_l,y_l)
\end{equation}
Notice how the method incorporates a denoising mechanism, so
no smoothing procedures are needed.
The great advantage of this method is that is possible to obtain a reasonable
approximation to CSD even when the presence of unusable channels is near or inside
the region of interest. The difference method, on the other hand, affects all surrounding
channels to a failing channel, up to 8 of them, so it becomes unusable if enough
saturated electrodes are over the r.o.i.

@@

The attainment of $K$ and $\tilde{K}$ turns out to be computer-costly for high density MEAs, but paraletization, specially on
GPUs, solves this issue.

%% posiblemente quitar esto 
We have performed comparison of both methods over experimental data that has
no failing channels in order to adjust the parameters of the second method. We
have obtained consistent results between them.

@@

Center of Mass Analysis (CMA) has been used previously in order to track putative  ``trajectories'' of the activity or information across neural tissue or even the whole brain \cite{Chao05, Chao07, Manjarrez07, Manjarrez09}. In such works the chosen ``mass'' was an heuristic measure adecuate for representing the corse-grain picture of displacement of activity. These present the problem of justifing the choice made on practical grounds. For calculating a vectorial average, the adecuate wheight per vector is a density, a positive quantity that measures the concentration of something per space unit, so it is only adequate that we search for a natural density in the electrophysiological experiments. The Current Source Density fits the mathematical and interpretative requirements. We shall denote it in the following as $C$ (following \cite{Potworowski2011}).  This measure is usually presented in arbitrary units, which permits us to interpret it two-fold. One, directly from its definition, as the \emph{differential divergence} of the vectorial electric current density:
\begin{equation}
  C:=\nabla \cdot J
\end{equation}
Due to conservation of charges, this is the same as as the temporal variation of density of charge:
\begin{equation}
  C=\frac{\partial \rho}{\partial t}
\end{equation}
Current Source Density as a mathematical density (a quasi-probability density function)  allows  to rigorously apply the concepts of vector averaging. This use has been overlooked. This might have occurred because  samplings at lesser spatial resolution did not permit estimations of this quantity with enough precision to make reliable spatial implications, even with the use of sophisticated CSD estimation methods. In our case, we want to use the density as a mathematician would: a function which assigns weights to points in the space of interest and permits us to carry linear operations over them.  The space here is the sampling points of the electrodes at a given instant in time, and the weight, the CSD, is amount of interchange of charged ions from the inside to the outside of cellular membranes, per unit space.  This is an indicator of activity in the neurons, therefore the CSD is also, indirectly, a \emph{``density of local activity''}. Having both
the mathematician and electrophysiologicist interpretations presents we can
perform the following analysis with clear readings.

Apparently, CSD would not fit the requirements of non-negativity for its use as a mathematical density, but this is not a real problem.  CSD separates the measured activity into three distinct sets, namely, the set of all sources, the set of all sinks and the zero set. The separation is not quantitative but qualitative: even when the signs of electric charge are arbitrary, the distinction between them is not. Therefore, CSD represents on each of the two active sets (sinks and sources) exactly what we need, a unit of concentration of changing charges per unit of space (in this case, per area), and the sign can be ignored, as long as we perform the operations in each set separately. Also, the separation by the zero value set of the CSD is in contrast to the zero value of the LFP, where it has no precise meaning as it only represents a practical reference value. This is acknowledged in most of the literature, where only the scales of the LFP recording are shown, without reference to the sign of the values. In the CSD representation,  the so called zero set (all the points in our space which have a recorded CSD of zero)  has a precise meaning: it is an instantaneous lack of activity and the border between sinks and sources. This shall be exploited further in the section below.

Once the separation into sinks and sources has been made, we would have two densities, or more precisely, two different density functions defined over two separate sets. Trying to use this for obtaining vector averages would still be crude and inexact. If these sets spread out over large non-convex areas, the vector averages could lie outside the sets, having little or none physiological interpretation. Therefore we decided to introduce another concept from geometrical sets in order to reduce the analysis to local effects.

\subsection{CSD and Disconnected Components}

A first visual inspection of the CSD color map at a given instant in time shows that both sources and sinks appear in rather large and significant patches. These appear to correspond to the expected double and triple poles of firing neurons in structured tissue \cite{Buzsaki2012} and inspired us to separate the sink and source  sets into their \emph{disconnected components}. This concept is used in mathematics to denote subsets of a given geometric set that do not touch each other  \cite{Halmos}. Their rough correspondence to poles of firing neurons would be most useful in tracking the successive locus of the activity.

To clarify the concept and how we are we using it an example is in order. In figure \ref{disconnectedsets} A we show geometrical sets enclosed by simple curves as color patches, where every color indicates a Set. The set $A$ consists of a single connected component, that means that one can make  path between any two points in the set that consist entirely of points of the same set. In contrast, the set $B$ is made of two disconnected components, labeled $B1$ and $B2$. If one chooses a point in each of the components, one cannot draw a path joining them without leaving the set. But each one of those components is connected, meaning that if we choose the pair of points in one component, then again we can trace a path joining them in the set. So the set $B$ consists of two disconnected components, but each component can be regarded as a connected set on their own. Likewise the set $C$ is made of three disconnected components. In the subfigure \ref{disconnectedsets} Aiii we show the center of mass of the sets. The set $A$ is concave, so, it can have a center of mass outside the set. The
set $B$ has also a center of mass outside the set. It has two disconnected components that are aproximatelly
equal and separated by a long distance (this is not an implication: a set can be disjoint and have
a CM inside the set). For the set $C$ our aproach has been used: we treat every disjoint component of the
set as an independent entity and calculate the center of mass for each one. Notice that this doesn't preclude the posibility of having a concave subset with a CM that lies outside. But, as we shall see,
as the electrode resolution is finite, the posibility of having highly non compact shapes is unlikely, and they
tend to have a CM inside or near their borders. In the figure \ref{disconnectedsets} B we show this with experimental
data (we are using evoked activity data, as explained in section \ref{sec:evocada}). In graph Bi we
sepparate the CSD data into the sink and sources set. We take sinks as example. In the graph Bii
we label each disjoint component, obtained by a single pass algorhythm. There are 25 components, and
the biggest two are not strictly convex but they have tolerable deviations from it. In the
subfigure Biii we indicate the CM of only those components that are above certain threshold of
integrated intensity (the dot is proportional in area to such intensity, there are five CM to be
seen). Each CM gets listed as a probable step in a trajectory.

\begin{figure}
  \includegraphics[width=0.9\textwidth]{Figuras/Disjuntos01.pdf}
  \caption{ Explanation of the geometrical concepts used on our analysis. In subfigure A we show
    with an artificial example the concepts of connectedness and convexivity. The subfigure Ai shows
    three sets, $A$, $B$, and $C$. $A$ fails to be convex because it is concave, and $C$ fails to be
    be convex because it is made of two disconnected components. In each of them is possible to draw a line
    connecting two points inside the set that is not entirely inside the set. In $B$ that is not possible, the set $B$ is convex in a mathematical sense. In subfigure Aii we made some modification to show that
    each set can be made of many disconnected components, and each one of them can be convex or not.  In Aiii we show the center of mass for $A$ and $B$ taken as a whole, while for $C$ we break the set into each of the disjoint components and calculate each CM separately. There is no guarantee that each disconnected component would be convex, but as the resolution of the data is finite,  small components are more likely to
    be convex. In figure B we show everything with a frame of experimental data. In Bi we sepparate the
    data into three sets, Sources, Sinks, and Zeros. We discard the Zeros and keep the active ones.
    In Figure Bii we take the Sink Set and separate it into disconnected components,
    they are labeled from one to 25 in this particular example.
    From this subsets we calculate the total mass of each one and only
    considerate for the next step those who are above certain threshold of integrated intensity. The
    next step is shown in Biii, where we plot the Center of Mass for each of such subsets. Deviations
    from convexity seem to be unimportant, as most of the mass is concentrated in the inner electrodes
    of each set.
  }
  \label{disconnectedsets}
\end{figure}

@@

The ``thick zero'' would help us to sepparate the sinks and sources sets into its disconnected components, which would correspond then to clear patches of definite activity. If each one of these patches could be interpreted as an active unit, we may asign to it a ``center of activity'' in order to follow their displacements. A Center of Mass for each instantaneous disconnected sink/source would provide us a pragmatic measurement of an ``active locus''. We can then calculate at each time frame the CM for each one of those components before tracking displacements of activity.

At each component at a precise time $t$, indexed $k$, we have a number of points with CSD of the same sign, indexed $j$. Then, the CM of the $k$ component would have the usual definition:
\begin{equation}\label{cmparadisj}
   \langle q(t) \rangle_k =\frac{\sum_j q_{j,k} (t) I_{j,k} (q_{j,k},t)}
           {\sum_j I(q_{j,k},t)},
\end{equation}
where $I(q_{j,k},t)$ is the CSD at point $q_{j,k}=(x_{j,k}, y_{j,k})$ at time $t$, and the
sign of our ``mass'' eliminates itself graciously.
These disconnected components, while not rigorously convex, are less spread over the space of the measurements
and tend to have their CM over the actual active points, or at the border of the active channels.

@@

We must not take out of sight the phenomena which is producing the CSD distribution: neurons have a finite size, and the effect of their action a delimited extent. Every patch that we can identify as a disconnected component of the Sources/Sink sets indicates a  joint active set of units. A center of mass for each disjoint component would then indicate an average locus for this active group, a putative center of activity to which we can give a position, intensity and time. An instantaneous snapshot of such information would depict the centers of activity of an ordered structure of neurons and their relative intensities. But it is in the successive depiction of a series of such snapshots that this analysis can show interpretative power. The appearance , displacement, and disappearance of such centers could reveal very subtle details of the dynamics of neural activity. Very small, fast displacements of the Centers of mass, may be occur due to slight asynchronous between neurons which belong to the same active group. Larger displacements and  longer intervals (more than two or so firing periods) could indicate physiological connections between different groups or units.

The next step in our analysis is a procedure to systematically associate to each CM at time $t$ a ``successor'' at time $t+\Delta t$, where $\Delta t$ is the sampling time step. Let $q_1=(x_1, y_1)$ be the coordinates of the first point and $q_2=(x_2,y_2)$, the corresponding coordinates of the second. Remember that these coordinates can be now rational numbers
respect to the grid of electrodes, as CM are not restricted to be  at the grid points.
The distance is:
\begin{equation}
d(q_1,q_2):=\sqrt{(x_1-x_2)^2+(y_1+y_2)^2}.
\end{equation}
The integrated intensity of a CM is simply the denominator in the eq. \ref{cmparadisj}, that is, the total ``mass'' obtained by integrating the CSD in a specific component:
\begin{equation}
  I_k(t)=\sum_j I (q_{j,k}, t).
\end{equation}
We specify then a tolerance $\delta$. If a certain $t$ a CM is below $\delta$ distance from another at $t+\Delta t$, we consider the latter the successor in time of the former. By continuously applying such procedure, we can begin to trace trajectories of the CMs, as they appear, wander, and fade.

When a specific trajectory cannot find a successor point that is above the
threshold intensity, we assume that it has ended, so we do not concatenate
points that are separated in time by more than $\Delta t$. In order to paste different
trajectories, more rigorous analysis of their origin and interpretation should be used and is
beyond the scope of this work.

@@

@@

\subsection{Counterexample: the stratial neuronal tissue}

bla bla.
